### 2014-05-13 tex/main-body-text.tex
Write technical description of flock
Also some minor edits to eric's earlier sections
and I made a jpg out of Patrick's big figure for
previewing, just because my computer can't handle the big
PDF file elegantly enough to have it there during
editing.

\subsection*{{\sc structure}}
In the {\sc structure} model with no admixture, the unknown ``subpopulation'' that the 
$i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, 
where $K$ is the 
number of subpopulations (or clusters, as they are often referred to).  
In a diploid individual from cluster $k$, the 
allelic types of the two gene copies at the $\ell\thh$ locus are assumed
to be drawn independently from the vector of allele frequencies at
locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where 
$A_\ell$ 
denotes the number of alleles observed in the data set at locus $\ell$.
Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the 
number
of copies of each of the $A_\ell$ alleles at locus $j$ in individual $i$, and 
$Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
$A_\ell$  components and cell probabilities given by the allele frequencies: 
\begin{equation}
(Y_{i\ell}~|~Z_i=k) \sim \mathrm{Mult}_{A_\ell}(2, \theta_{k\ell}).
\end{equation}
In the model without physical linkage, the genotypes at the loci are assumed to
be independent of one another, so the probability of the genotype data at
all $L$ loci---$Y_i=(Y_{i1},\ldots,Y_{iL})$---is simply a product of multinomial 
probabilities.
In the uncorrelated allele frequencies model, the prior on each $\theta_{k\ell}$ is a
Dirichlet distribution with parameters $(\lambda_{k\ell1},\ldots,\lambda_{k\ell A_
\ell})$,
which are usually set to a value like $1/A_\ell$.

Inference in the model proceeds by sampling from the joint posterior of the 
$Z_i$'s and $\theta = (\theta_1,\ldots,\theta_K)$ using Gibbs sampling.  That is,
at iteration $t$:
\begin{enumerate}
\item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
the full conditional distribution of $Z_i$ given
$Y_i$ and $\theta^{(t)}$ (the current estimate of the allele frequencies).  This
distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each 
cluster
$k$ and locus $\ell$, the full conditional for $\theta_{k\ell}$ is independently a 
Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
$j=1,\ldots, A_\ell$,
where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
individuals with $Z^{(t+1)}_i = k$.   
\end{enumerate}

It will be useful for our comparison with {\sc flock} to point out that another way 
of pursuing inference in this model would be to first integrate out the 
Dirichlet priors on the allele frequencies and then sample from the posterior
for the $Z_i$'s by Gibbs sampling.  When $\theta$ is integrated out, the genotypes 
of the individuals are no longer conditionally independent (they were originally
independent {\em conditional} on $\theta$ and the $Z_i$'s), so the calculation of 
the full conditional distribution of $Z_i$ becomes somewhat more involved,
but can be computed by the following reasoning. First, the full 
conditional distribution of $Y_{i\ell}$, given that individual $i$ is from subpopulation $k$
(\ie $Z_i=k$) depends on the 
cluster memberships and genotypes  of all the remaining individuals,
which we denote by  $Z_{(-i)}$ and $Y_{(-i)\ell}$, respectively.
This full conditional for $Y_{i\ell}$ follows 
a Compound Dirichlet Multinomial distribution (CDM):
\begin{eqnarray}
\lefteqn{(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell}) \sim} \label{eq:cdm}\\
& & \mathrm{CDM}(\lambda_{k\ell 1} + \#_{(-i)}(k,\ell,1), \ldots,
\lambda_{k\ell A_\ell} + \#_{(-i)}(k,\ell,A\ell)), \nonumber
\end{eqnarray}
where $\#_{(-i)}(k,\ell,1)$ is the number of alleles of type $j$ at locus $\ell$
found in individuals {\em other than individual $i$} that currently
belong to cluster $k$. Thus, calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
\{1,\ldots,K\}$ 
and normalizing to sum to one gives the full conditional distribution for 
$Z_i$:
\begin{equation}
P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})~~,~~k=1,\ldots,K.
\label{eq:fc}
\end{equation}
A new value of $Z_i$ would be drawn from (\ref{eq:fc}) if doing Gibbs sampling in this
version of the {\sc structure} model in which $\theta$ has been integrated out.  In 
fact,
this is the approach (with a slightly different prior weight on the $Z_i$'s) taken to 
update 
$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} 
when not
proposing changes to the number of clusters.

@@

The first step of the flock algorithm is initialization, during which 
the program randomly allocates each of the $N$ individuals to one of 
$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is given the chance to be reallocated 
(\ie moved to a different cluster).  This is done on the basis of 
maximum likelihood:

One issue I have with it is that it is described as a ``procedure'' or 
an ``algorithm,'' without an attempt to see it as a special case of 
well known inference frameworks.

\subsection*{{\sc flock} as a special case of {\sc structure}}

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.pdf}
    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 
in the program {\sc flock}. Each horizontal plot represents the 'best run' from a 
series where individuals are reallocated into \textit{k} clusters. Within each plot 
the normalized individual likelihood for each reference population is represented by a 
colored vertical bar. Major geographic boundaries that were attributed to genetic 
structure in \citet{Garzaetal_norcal} are indicated with black vertical lines. Results 
from the 'best runs' showed similar geographic differentiation to 
\citet{Garzaetal_norcal} analysis. The plateau analysis developed by 
\citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 
is the lower bound estimate.}
    \label{Fig.1}
\end{center}
\end{figure*}

>>>>

\subsection*{{\sc structure}}
In the {\sc structure} model with no admixture, the unknown ``subpopulation'' that the 
$i\thh$ individual ($i=1,\ldots,N)$ belongs to is denoted by $Z_i \in \{1,\ldots,K\}$, 
where $K$ is the 
number of subpopulations (or clusters, as they are often referred to).  
In a diploid individual from cluster $k$, the 
allelic types of the two gene copies at the $\ell\thh$ locus are assumed
to be drawn independently from the vector of allele frequencies at
locus $\ell$,  $\theta_{k\ell}=(\theta_{k\ell 1},\ldots,\theta_{k\ell A_\ell})$, where 
$A_\ell$ 
denotes the number of alleles observed in the data set at locus $\ell$.
Hence, if $Y_{ij}$ denotes a vector of length $A_\ell$ whose components denote the 
number
of copies of each of the $A_\ell$ alleles at locus $j$ in individual $i$, and 
$Z_i=k$, then $Y_{i\ell}$ follows the multinomial distribution of two trials with
$A_\ell$  components and cell probabilities given by the allele frequencies: 
\begin{equation}
(Y_{i\ell}~|~Z_i=k) \sim \mathrm{Mult}_{A_\ell}(2, \theta_{k\ell}).
\end{equation}
In the {\sc structure} model without physical linkage, the genotypes at the loci are assumed to
be independent of one another, so the probability of the genotype data at
all $L$ loci---$Y_i=(Y_{i1},\ldots,Y_{iL})$---is simply a product of multinomial 
probabilities.
In the uncorrelated allele frequencies model, the prior on each $\theta_{k\ell}$ is a
Dirichlet distribution with parameters $(\lambda_{k\ell1},\ldots,\lambda_{k\ell A_
\ell})$,
which are usually set to a value like $1/A_\ell$.

After initializing the unknown allele frequencies to randomly drawn values,
$\theta^{(0)}$, inference in the model proceeds by sampling from the joint posterior of the 
$Z_i$'s and $\theta = (\theta_1,\ldots,\theta_K)$ using Gibbs sampling.  That is,
at iteration $t = 0, 1, 2, \ldots$:
\begin{enumerate}
\item each $Z^{(t)}_i$ is updated to $Z^{(t+1)}_i$ by sampling a value from 
the full conditional distribution of $Z_i$ given
$Y_i$ and $\theta^{(t)}$ (the current estimate of the allele frequencies).  This
distribution is found using Bayes' theorem. In {\sc structure}'s formulation, the 
prior probability that $Z_i=k$ is $1/K$ for all $k=1,\ldots, K$.     
\item $\theta^{(t)}$ is updated from its full conditional distribution.  For each 
cluster
$k$ and locus $\ell$, the full conditional for $\theta_{k\ell}$ is independently a 
Dirichlet distribution with parameters $\lambda_{k\ell j} + \#(k,\ell,j)$, for 
$j=1,\ldots, A_\ell$,
where $\#(k,\ell,j)$ is the number of alleles of type $j$ at locus $\ell$ observed in 
individuals with $Z^{(t+1)}_i = k$.   
\end{enumerate}

It will be useful for our comparison with {\sc flock} to point out that another way 
of pursuing inference in this model would be to first integrate out the 
Dirichlet priors on the allele frequencies and then sample from the posterior
for the $Z_i$'s by Gibbs sampling.  When $\theta$ is integrated out, the genotypes 
of the individuals are no longer conditionally independent (they were originally
independent {\em conditional} on $\theta$ and the $Z_i$'s), so the calculation of 
the full conditional distribution of $Z_i$ becomes somewhat more involved,
but can be computed by the following reasoning. First, the full 
conditional distribution of $Y_{i\ell}$, given that individual $i$ is from subpopulation $k$
(\ie $Z_i=k$), depends on the 
cluster memberships and genotypes  of all the remaining individuals,
which we denote by  $Z_{(-i)}$ and $Y_{(-i)\ell}$, respectively.
This full conditional for $Y_{i\ell}$ follows 
a Compound Dirichlet Multinomial distribution (CDM):
\begin{eqnarray}
\lefteqn{(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell}) \sim} \label{eq:cdm}\\
& & \mathrm{CDM}(\lambda_{k\ell 1} + \#_{(-i)}(k,\ell,1), \ldots,
\lambda_{k\ell A_\ell} + \#_{(-i)}(k,\ell,A\ell)), \nonumber
\end{eqnarray}
where $\#_{(-i)}(k,\ell,1)$ is the number of alleles of type $j$ at locus $\ell$
found in individuals {\em other than individual $i$} that currently
belong to cluster $k$. Thus, calculating (\ref{eq:cdm}) for each value of $Z_i=k \in 
\{1,\ldots,K\}$ 
and normalizing to sum to one gives the full conditional distribution for 
$Z_i$:
\begin{equation}
P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})~~,~~k=1,\ldots,K.
\label{eq:fc}
\end{equation}
A new value of $Z_i$ would be drawn from (\ref{eq:fc}) if doing Gibbs sampling in this
version of the {\sc structure} model in which $\theta$ has been integrated out.  In 
fact,
this is the approach (with a slightly different prior weight on the $Z_i$'s) taken to 
update 
$Z_i$ in both {\sc hwler} \citep{Pel&Mas2006} and {\sc structurama} \cite{Hue&And2007} 
when not
proposing changes to the number of clusters.

@@

The first step of the flock algorithm is initialization, during which 
the program randomly allocates each of the $N$ individuals to one of 
$K$ clusters.  Then a number of reallocation steps are performed.  During every one of 
these steps, each individual is given the chance to be reallocated 
(\ie moved to a different cluster).  This is done on the basis of 
maximum likelihood: as \citeauthor{Duc&Tur2009} say (2009, p.~1335), ``Re-allocations are 
performed following multilocus maximum likelihood (Paetkau et al. 1995).''

From that description, it is not clear how \citeauthor{Duc&Tur2009} treat 
alleles that appear in an
individual but do not appear within any other individuals within a cluster.
The original approach of \citet{Paetkauetal1995} merely added 0.01 or some other
small value to each allele frequency that was 0, and then renormalized 
the allele frequencies to sum to 1.  In later works, Paetkau and colleagues
also employed the ``Bayesian'' approach of \citet{Ran&Mou1997} in their 
software {\sc geneclass2}  \citep{Piryetal2004}.  If \citeauthor{Duc&Tur2009}
use the approach of \citet{Ran&Mou1997}, then the $i\thh$ individual will
be reallocated to whichever cluster gives the highest value to 
$P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$, which is exactly the probability
defined in (\ref{eq:cdm}).  
If \citeauthor{Duc&Tur2009} use the simpler ``add 0.01 and renormalize''
formulation of \citet{Paetkauetal1995}, then their reallocations are based
on maximizing a likelihood that, although it may not be formally identical to
that in (\ref{eq:cdm}), is not substantively different from it.

It is worth pointing out that, since the
{\sc structure} model without admixture assumes a uniform prior over 
the $K$ different clusters, $P(Y_{i\ell}~|~Z_i=k,~Z_{(-i)},~Y_{(-i)\ell})$
is exactly proportional to $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$, which 
we have shown is the
full conditional distribution for $Z_i$ in the {\sc structure} model after
integrating out $\theta$.  Thus, we have shown that, doing Gibbs
sampling for $Z_i$ in  the {\sc structure} in which $\theta$ has been
integrated out would involve sampling a new value of $Z_i$ from the 
probability distribution,
\[
P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
\]
and that the updates that
{\sc flock} makes to each $Z_i$ involve assigning to $Z_i$
\begin{equation}
\arg\max_k P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})
\end{equation}
\ie, the value of $k$ that maximizes $P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})$
(or, as stated above, a very similar function of $k$).

\subsection*{{\sc flock} as a special case of {\sc structure}}
The goal of {\sc structure}'s MCMC algorithm is to sample from the space of 
$Z$'s in proportion to their posterior probability.  If, all that was desired
was a point estimate of the $Z_i$'s that maximized the posterior probability,
then one standard approach to finding that maximimum {\em a posteriori}, or MAP, 
estimate, would be to use simulated annealing \citep{Kirkpatricketal1983}.
To do simulated annealing in the {\sc structure} model with no admixture and
no correlated allele frequency prior, after integrating out the unknown
allele frequencies, the updates for each $Z_i$ would be made from the 
distribution
\[
\biggl[P(Z_i=k~|~Y_i, ~Z_{(-i)},~Y_{(-i)\ell})\biggr]^\beta
\]
which is the same full conditional distribution as before, raised
to the power $\beta$.  In typical applications of simulated annealing,
$\beta$ is initially set to a low value $<1$ and as the algorithm proceeds
$\beta$ is increased according to a ``cooling schedule''\citep{Hajek1988}, until
the algorithm finds a local maximum in the posterior probability surface.  
When $\beta$ is small, the process can easily traverse ``valleys''
in the posterior probability surface and have a better chance of searching
more of the space for the maximum value.  As $\beta$ gets larger, the process
tends to move uphill on the probability surface, searching for the
maximum near where it currently is.

It should now be clear that the algorithm in {\sc flock} is a simulated annealing 
algorithm for the {\sc structure} model in which the cooling schedule is
``start with $\beta\rightarrow\infty$ and leave it there for the duration of 
the algorithm.''  As such, we can expect that it may not provide the best 
mechanism for searching for the MAP estimate and it may be more susceptible 
to becoming trapped in local modes of the posterior probability. 
In the following section we show that {\sc structure} and {\sc flock} find similar
solutions in the analysis of a large microsatellite data set, though, as expected,
{\sc flock} seems to find the same solution less consistently than does {\sc structure}.

\subsection*{boneyard}
 On the other hand, in practice, the 
posterior probability surface may have several modes that are separated by regions
of low posterior probability, such that the solution
an algorithm converges to (whether it is {\sc structure}  doing MCMC or {\sc flock} doing 
``very cold simulated annealing'') depends almost entirely on the initial random
starting values.

@@

\begin{figure*}
\begin{center}
	 % NOTE. Eric made Flock-Fig1.jpg as a Preview because Flock-Fig1.pdf
	 % ate up too many resources while editing the document.  For final 
	 % production, it should be changed back to Flock-Fig1.pdf
    \includegraphics[width=\textwidth]{images/Figures-Pat/Flock-Fig1.jpg}  
    \caption{Assignment to reference population (from \textit{k} = 2-6) plots created 
in the program {\sc flock}. Each horizontal plot represents the 'best run' from a 
series where individuals are reallocated into \textit{k} clusters. Within each plot 
the normalized individual likelihood for each reference population is represented by a 
colored vertical bar. Major geographic boundaries that were attributed to genetic 
structure in \citet{Garzaetal_norcal} are indicated with black vertical lines. Results 
from the 'best runs' showed similar geographic differentiation to 
\citet{Garzaetal_norcal} analysis. The plateau analysis developed by 
\citet{Duc&Tur2012} indicate that either an undecided structure or that \textit{k}=2 
is the lower bound estimate.}
    \label{Fig.1}
\end{center}
\end{figure*}
